Synthesis of Linear Functional Programs

### 2021-05-09 article/article.tex
finish background

Richer type systems allow for more precise types, which can
statically eliminate various kinds of logical errors by making certain
invalid program states ill-typed (e.g.,~a ``null aware'' type system,
will ensure at compile-time that you cannot dereference a
null-pointer). However, they can also be a burden -- the whole point
of these type systems is to ensure that \emph{fewer} programs are
deemed well-typed, which can pose additional challenges to the
development process. Type-based synthesis leverages rich types as a
way of pruning the search space, and by using types gives the user as
a more ``familiar'' specification. For instance, the type
$\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow \mathsf{Int}$
specifies a (curried) function that takes two integers and produces an
integer. Viewed as a specification, it is extremely imprecise (there
are an infinite number of functions that satisfy this specification).
However, the richer type $(x{:}\mathsf{Int}) \rightarrow
(y{:}\mathsf{Int}) \rightarrow \{z{:}\mathsf{Int} \mid z = x+y\}$
very precisely specifies a function that takes integer arguments $x$
and $y$ and returns an integer $z$ that is the sum of $x$ and $y$.

@@

The Curry-Howard correspondence generalizes beyond the simply-typed
$\lambda$-calculus and propositional intuitionistic logic. It extends
to the realms of polymorphism (higher-order logic \todo{review -
  \url{https://en.wikipedia.org/wiki/Lambda_cube}} second-order
propositional calculus), dependent types (first-order predicate
logic), and various other extensions of natural deduction and
simply-typed $\lambda$-calculus -- which include linear logic and
linear types.

@@

  The elimination form for the exponential, written $\llet{ !u
    = M}{ N}$, warrants the use of resource $A$ an unbounded
  number of times in $N$ via the variable $u$:  
\[
    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta_1 \vdash M : \bang A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
    {\Gamma ; \Delta_1, \Delta_2 \vdash \llet{ !u = M}{ N} : C}
\]

\todo{Through the Curry-Howard correspondence, we can view the process
  of finding a proof of a proposition as the process of synthesizing a
program of a given type.}

@@

% \todo[inline]{Os~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
%   como ``estrategia geral'' de ``ler'' um sistema de tipos +
%   engenharia como um sistema de sintese de programas tipificados;
%   Aqui~\cite{DBLP:conf/pldi/PolikarpovaKS16} estamos num contexto
%   ``mais avançado'' (polimorfismo e tipos refinados -- ``logica de
%   primeira ordem'') -- tipos ricos -- trabalho aspiracional,
%   tratamento de funcao recursivas, tratamento de tipos de dados
%   indutivos/algebricos; Aqui~\cite{DBLP:conf/lopstr/HughesO20},
%   sintese para graded modal types, tem a ver com ``linearidade'' kind
%   of, abordagem baseado em focusing (portanto Curry-Howard++), no caso
%   deles acaba por ser uma lógica ``modal'' vs linear -- não permite
%   futuramente explorar relações com e.g. concorrência, análise de
%   recursos.}

@@

% (no recursion   -- in that work, no
%   obvious connection of grading with
%   concurrency)

% Refinement types provide very precise specifications to assist the
% synthesis process.

% \item Synthesis of heap-manipulating
%   programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}

% Less related, but more evidence of this idea of turning ``checking''
% systems into synthesis frameworks.

% \item Resource-guided synthesis \cite{DBLP:conf/pldi/KnothWP019}

% This resource-guided means something a bit different. Programs satisfy
% a functional specification and a symbolic resource bound in the sense
% of amortized analysis, but can provide technical insights.

@@

Since proof rules in \emph{natural deduction} do not entail a
proof-search strategy, we will use the equivalent \emph{sequent
  calculus} system to formulate our typing rules in a way that is
amenable to proof search.
%
In sequent calculus, rules can be naturally understood in a
\emph{bottom-up} manner, simplifying the reasoning for the
synthesizer. Moreover, the linear nature of our system requires a way
to algorithmically reason about resource contexts: for instance, if we
read rule $\tensor I$ from Section~\ref{sec:background} bottom-up, we
must non-deterministically guess how to correctly split ambient
resources into $\Delta_1$ and $\Delta_2$ to separately prove $A$ and
$B$. To solve this issue we will adopt the resource-management
approach
of~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:conf/lics/LiangM09}.
Instead of requiring non-deterministic guesses of resource splits
during proof search, we will make use of an \emph{input} and an
\emph{output} context, leading to the following general strategy: to
prove $A\tensor B$ using resources $\Delta$, prove $A$ with $\Delta$,
consuming some subset of $\Delta$, and produce as output the leftover
resources $\Delta'$; prove $B$ using $\Delta'$ and then output the
remaining resources $\Delta''$; finally, after having proven
$A\tensor B$, output $\Delta''$, for subsequent derivations.

@@

However, the search process still encompasses significant
non-determinism, e.g., at any given point, many proof rules are
applicable in general. To address this we draw on the technique of
focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} --
a method created to trim down the search space of valid proofs in
linear logic, by eagerly applying invertible rules (i.e.~rules whose
conclusion implies the premises), and then by ``focusing'' on a single
connective when no more direct (invertible) rules can be applied. If
the search is not successful, the procedure backtracks and another
connective is chosen as the focus.
%
Focusing eliminates all of the
``don't care'' non-determinism from proof search, since the order in
which invertible rules are applied does not affect the outcome of the
search, leaving only the non-determinism that pertains to unknowns (or
``don't know'' non-determinism), identifying precisely the points at
which backtracking search is necessary.

@@

The full and partial synthesis mentioned are closely related, in which the latter is also a complete synthesis, but inserted inside another, bigger, program that makes use pf the synthetized function.

\mypara{Work Plan}
The end goal is to have a synthetizer capable of taking a type (possibly refined) as input,
and to produce (possibly interactively) a function or more (possibly recursive) with that same type.
Besides the synth, we've already created a parser and a type-checker -- more than good preparation for the
development of the synthetizer, they will be useful to \emph{validate} the functions synthetized, and to implement the
synthesis of partial programs. The implementations are in Haskell, a functional language, since it's easier to reason about inference rules that way.

>>>>

Richer type systems allow for more precise types, which can
statically eliminate various kinds of logical errors by making certain
invalid program states ill-typed (e.g.,~a ``null aware'' type system,
will ensure at compile-time that you cannot dereference a
null-pointer). However, they can also be a burden -- the whole point
of these type systems is to ensure that \emph{fewer} programs are
deemed well-typed, which can pose additional challenges to the
development process. Type-based synthesis leverages rich types as a
way of pruning the search space, and by using types the user 
is given a more ``familiar'' specification. For instance, the type
$\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow \mathsf{Int}$
specifies a (curried) function that takes two integers and produces an
integer. Viewed as a specification, it is extremely imprecise (there
are an infinite number of functions that satisfy this specification).
However, the richer type $(x{:}\mathsf{Int}) \rightarrow
(y{:}\mathsf{Int}) \rightarrow \{z{:}\mathsf{Int} \mid z = x+y\}$
very precisely specifies a function that takes integer arguments $x$
and $y$ and returns an integer $z$ that is the sum of $x$ and $y$.

@@

The Curry-Howard correspondence generalizes beyond the simply-typed
$\lambda$-calculus and propositional intuitionistic logic. It extends
to the realms of polymorphism (second-order
logic), dependent types (first-order
logic), and various other extensions of natural deduction and
simply-typed $\lambda$-calculus -- which include linear logic and
linear types.

@@

The elimination form for the exponential, written $\llet{ !u
= M}{ N}$, warrants the use of resource $A$ an unbounded
number of times in $N$ via the variable $u$:  
\[
    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta_1 \vdash M : \bang A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
    {\Gamma ; \Delta_1, \Delta_2 \vdash \llet{ !u = M}{ N} : C}
\]
Again, through the Curry-Howard correspondence, we can view the process
of finding a proof of a proposition in this linear logic as the process of synthesizing a
linear functional program of a given type.

\mypara{Resource-Management}
Coming from linear logic, we now have a new problem to solve:
When searching for a proof, how do we break the
contexts in the necessary way to allocate the correct resources
to prove each judgment the judgment we're proving depends on?
This is where the idea of \emph{input/output} contexts come in.
To solve this issue we will adopt the resource-management
approach
of~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:conf/lics/LiangM09}.
Instead of requiring non-deterministic guesses of resource splits
during proof search, we will make use of an \emph{input} and an
\emph{output} context, leading to the following general strategy: to
prove $A\tensor B$ using resources $\Delta$, prove $A$ with $\Delta$,
consuming some subset of $\Delta$, and produce as output the leftover
resources $\Delta'$; prove $B$ using $\Delta'$ and then output the
remaining resources $\Delta''$; finally, after having proven
$A\tensor B$, output $\Delta''$, for subsequent derivations.

\mypara{Sequent Calculus}
Inference rules in natural deduction (the ones we have considered so far)
are complicated to use in a proof-search strategy.
A more suited candidate is the equivalent \emph{sequent calculus} system
in which \emph{all} inference rules are read naturally in a \emph{bottom-up} manner.
The introduction and elimination rules from natural deduction disappear,
and its place is taken by right and left rules, respectively.
Right rules and introduction rules are almost identical, since introduction rules were
already read bottom-up. The inference rule for implication introduction ($\supset$ is used instead of $\rightarrow$), seen above in natural deduction under the
\emph{propositions as types} subsection, is the following right rule in sequent calculus:
\[
    \infer*[right=($\supset R$)]
    {\Gamma, u: A \vdash M : B}
    {\Gamma \vdash \lambda u. M : A \supset B}
\]
Left rules try to capture the same idea as elimination rules in natural deduction, but are altered
to reflect it bottom-up, instead of top-down (usual ``orientation'' of elimination rules).
The inference rule for the also seen above implication elimination, is the following left rule here:
\[
    \infer*[right=($\supset L$)]
    {\Gamma , u: A \supset B \vdash M : A \supset B \and \Gamma \vdash N : A}
    {\Gamma \vdash M\,N : B}
\]
We'll define our inference rules for synthesis under the sequent calculus system,
for it simplifies the work to be done in the synthetizer.

\mypara{Focusing}
Even with everything mentioned so far, non-determinism still pertains proof-search in logic, e,g., at any given point, many proof rules are applicable in general.
Focusing has been previously studied as a way to conduct proof search in linear logic~\cite{}
--
a method created to trim down the search space of valid proofs in
linear logic, by eagerly applying invertible rules (i.e.~rules whose
conclusion implies the premises), and then by ``focusing'' on a single
connective when no more direct (invertible) rules can be applied. If
the search is not successful, the procedure backtracks and another
connective is chosen as the focus.
%
Focusing eliminates all of the
``don't care'' non-determinism from proof search, since the order in
which invertible rules are applied does not affect the outcome of the
search, leaving only the non-determinism that pertains to unknowns (or
``don't know'' non-determinism), identifying precisely the points at
which backtracking search is necessary.

@@

@@

% \item Resource-guided synthesis \cite{DBLP:conf/pldi/KnothWP019}
% This resource-guided means something a bit different. Programs satisfy
% a functional specification and a symbolic resource bound in the sense
% of amortized analysis, but can provide technical insights.

@@

Since proof rules in \emph{natural deduction} do not entail a
proof-search strategy, we will use the equivalent \emph{sequent
calculus} system to formulate our typing rules in a way that is
amenable to proof search.
%
In sequent calculus, rules can be naturally understood in a
\emph{bottom-up} manner, simplifying the reasoning for the
synthesizer. Moreover, the linear nature of our system requires a way
to algorithmically reason about resource contexts: for instance, if we
read rule $\tensor I$ from Section~\ref{sec:background} bottom-up, we
must non-deterministically guess how to correctly split ambient
resources into $\Delta_1$ and $\Delta_2$ to separately prove $A$ and
$B$. The resource-management approach we'll adopt to solve this has been 
previously explained in more detail.

@@

However, the search process still encompasses significant
non-determinism. To address this we draw on the technique of
focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} as seen before.

@@

\mypara{Work Plan}
The end goal is to have a synthetizer capable of taking a type (possibly refined) as input,
and to produce (possibly interactively) a function or more (possibly recursive) with that same type.
We've already created a parser and a type-checker -- more than good preparation for the
development of the synthetizer, they will be useful to \emph{validate} the functions synthetized, and to implement the
synthesis of partial programs. The implementations are in Haskell, a functional language, since it's easier to reason about inference rules that way.
